Bayesian Neural Arithmetic Units -- Leveraging The Complex Plane

### 2020-04-22 paper/main.tex

It is known that Neural Networks can approximate functions arbitrarily
well\todo{cite universal approx}.  However, this strength of approximation
often comes at the cost of generalization beyond the training data manifold.
\citet{trask_neural_2018} have demonstrated how severe this problem is even for
the simplest arithmetic operations, such as summing or multiplying two numbers.
In order to increase the power of abstraction of Neural Networks they propose a
\emph{Neural Arithmetic Logic Unit} (NALU) which can help to overcome this problem
\cite{trask_neural_2018} and is capable of learning the arithmetic addition
$+$, subtraction $-$, multiplication $\times$, and division
$\{+,-,\times,\div\}$ with stunning extrapolation accuracy.
However, the NALU comes with the severe limitation not being able to handle negative
inputs due to the logarithm in the multiplication part of the NALU:
\begin{align}
  \label{eq:nalu}
  \text{Addition: }       & \bm a = \bm W \bm x                  
                          & \bm W& = \tanh(\hat{\bm W}) \odot \sigma(\hat{\bm M}) \\
  \text{Multiplication: } & \bm m = \exp \bm W(\log(|\bm x|+\epsilon)) & &\\
  \text{Output: }         & \bm y = \bm a \odot \bm g + \bm m \odot (1-\bm g) 
                          & \bm g& = \sigma(\bm G\bm x)
\end{align}

A multiplication layer that can handle negative inputs was introduced by \citet{madsen_neural_2020}.
The \emph{Neural Multiplication Unit} (NMU) is defined by
\begin{align}
  \text{NMU: } y_j = \prod_i W_{ij} z_{i} + 1 - W_{ij} \hskip .2\textwidth W_{ij}=\min(\max(W_{ij}, 0), 1)
\end{align}

@@

We fuse the ideas of the \emph{Neural Multiplication Unit} (NMU) by
\citet{madsen_neural_2020} and the \emph{Neural Arithmetic Logic Unit} (NALU)
by \citet{trask_neural_2018}.

The NALU in its original form is capable of addition, subtraction, and
multiplication of \textbf{positive}, high-dimensional vectors. However, for
small inputs convergence is complicated by the use of the logarithm in the
NALU. The logarithm further prevents the NALU of processing negative inputs
correctly.

The NMU fixes both the previous issues, but can only multiply two inputs, and
not represent arbitrary power functions.  Addition is taken care of by the
\emph{Neural Addition Unit} (NAU), which is essentially a dense layer without
bias. A chain of NAU and NMU can thus learn the task defined in
Eq.~\ref{eq:nmu_task}.

\begin{equation}
  \label{eq:nmu_task}
  y = (x_1 + x_2) \cdot (x_1 + x_2 + x_3 + x_4)
\end{equation}

\section{Complex Neural Multiplication Unit}%
\label{sec:complex_neural_multiplication_unit}

Our \emph{complex NMU} (NMUX) can learn arbitrary power functions while still
being able to correctly deal with negative inputs.
The NMUX layer is defined by
\begin{align}
  k_i &= \begin{cases}
     0  & x_i \leq 0 \\
    \pi & x_i > 0
  \end{cases} \\
  \bm r &= |\bm x| \\
  \bm m &= \exp(\bm M \log(\bm r)) \odot \cos(\bm M \bm k)
\end{align}
where $\bm k$ is a vector that is zero where $\bm x$ is positive and $\pi$
where it is negative, $\bm r$ is the elementwise absolute value, and $\bm M$
the multiplication matrix that we want to learn.

The NMUX is inspired by the multiplication part of the NALU, extended to
negative inputs by leveraging taking advantage of the complex logarithm.
The multiplication part of the NALU reads:
\begin{equation}
  \bm m = \exp(\bm M \log(|\bm x| + \epsilon)),
\end{equation}
where $\epsilon$ is a small constant that ensures positive inputs to the real
logarithm.  If use the complex logarithm, this constant becomes superfluous,
but we end up with a complex output in case of negative $x_i$.  In general, for
any complex number $z$\footnote{Note that complex numbers can be represented in
the polar, complex plane where $z=re^{i\varphi}$ with $r>0$ and $\varphi \in
(0,2\pi)$.}
\begin{align}
  \log(z) = \log\left(r\cdot e^{i\varphi}\right)
     = \log(r) + i\varphi.
\end{align}
However, as $z$ is assumed to be real, we can simplify as follows:
\begin{align}
  \log(z) = \log(r) + ik\pi,
\end{align}
where $k \in [0,1]$.

>>>>

It is known that Neural Networks can approximate functions arbitrarily
well\todo{cite universal approx}.  However, this strength of approximation
often comes at the cost of generalization beyond the training data manifold.
\todo{mention ODEs as use case}

\section{Related Work}%
\label{sec:related_work}

\citet{trask_neural_2018} have demonstrated how severe this problem is even for
the simplest arithmetic operations, such as summing or multiplying two numbers.
In order to increase the power of abstraction of Neural Networks they propose a
\emph{Neural Arithmetic Logic Unit} (NALU) which can help to overcome this
problem \cite{trask_neural_2018} and is capable of learning the arithmetic
addition, subtraction, multiplication, and division
$\{+,-,\times,\div\}$ with stunning extrapolation accuracy.  However,
the NALU comes with the severe limitation not being able to handle negative
inputs due to the logarithm in the multiplication part of the NALU:

\begin{align}
  \label{eq:nalu_add}
  \text{Addition: }       & \bm a = \bm W \bm x
                          & \bm W& = \tanh(\hat{\bm W}) \odot \sigma(\hat{\bm M}) \\
  \label{eq:nalu_mult}
  \text{Multiplication: } & \bm m = \exp \bm W(\log(|\bm x|+\epsilon)) & &\\
  \text{Output: }         & \bm y = \bm a \odot \bm g + \bm m \odot (1-\bm g) 
                          & \bm g& = \sigma(\bm G\bm x)
\end{align}

A multiplication layer that can handle negative inputs was introduced by \citet{madsen_neural_2020}.
The \emph{Neural Multiplication Unit} (NMU) is defined by Eq.~\ref{eq:nmu} and is typically used in
conjunction with the so-called \emph{Neural Addition Unit} (NAU) in Eq.~\ref{eq:nau}.
\begin{align}
  \label{eq:nmu}
  \text{NMU: } &y_j = \prod_i M_{ij} z_{i} + 1 - M_{ij}  &M_{ij}=\min(\max(M_{ij}, 0), 1)\\
  \label{eq:nau}
  \text{NAU: } &\bm y = \bm A \bm x &A_{ij}=\min(\max(A_{ij}, 0), 1)
\end{align}
In both NMU and NAU the weights are clipped to $[0,1]$, and typically regularized
with $\mathcal{R}$:
\begin{align}
  \label{eq:rsparse}
  \mathcal{R} = \sum_{ij} \min(W_{ij}, 1-W_{ij})
\end{align}

The combination of NAU and NMU can thus learn $\{+,-,\times\}$, but no division.

@@

\section{Neural Power Unit}%
\label{sec:neural_power_unit}

Our \emph{Neural Power Unit} (NPU) can learn arbitrary power functions (which
includes division) while still being able to correctly deal with negative
inputs.  The NPU layer is defined by
\begin{align}
  k_i &= \begin{cases}
     0  & x_i \leq 0 \\
    \pi & x_i > 0
  \end{cases} \\
  \bm r &= |\bm x| \\
  \bm m &= \exp(\bm M \log(\bm r)) \odot \cos(\bm M \bm k)
\end{align}
where $\bm k$ is a vector that is zero where $\bm x$ is positive and $\pi$
where it is negative, $\bm r$ is the elementwise absolute value, and $\bm M$
the multiplication matrix that we want to learn.

The NPU is inspired by the multiplication part of the NALU
Eq.~\ref{eq:nalu_mult}, extended to negative inputs by taking advantage of the
complex logarithm. In general, for any complex number $z$\footnote{Note that
complex numbers can be represented in the polar, complex plane where
$z=re^{i\varphi}$ with $r>0$ and $\varphi \in (0,2\pi)$.}
\begin{align}
  \log(z) = \log\left(r\cdot e^{i\varphi}\right)
     = \log(r) + i\varphi.
\end{align}
However, as $z$ is assumed to be real, we can simplify as follows:
\begin{align}
  \log(z) = \log(r) + ik\pi,
\end{align}
where $k \in [0,1]$ is zero for positive inputs and one for negative inputs $z$.
